---
title: "218X R9 - San Francisco Burden and Parcel Dashboard"
output: flexdashboard::flex_dashboard
---


```{r global, include = F}
knitr::opts_chunk$set(warning = F, message = F)
library(tidyverse)
library(tigris)
library(sf)
library(mapview)
library(leaflet)
library(plotly)
library(censusapi)
library(mapboxapi)
library(readxl)
library(tidycensus)
library(flexdashboard)


Sys.setenv(CENSUS_KEY="0b9dfaf3fa0df7bb9bce2b33001ae985230cc118")

ca_pumsR6 <- readRDS("ca_pumsR6.rds")
sf_burden_pumas <- readRDS("sf_burden_pumas.rds")
my_parcel_zoning <- readRDS("my_parcel_zoning.rds")

burden_palOwner <- colorNumeric(
  palette = "Purples",
  domain = sf_burden_pumas %>% 
    filter(
      status == "Owner"
    ) %>% 
    pull(burdened_30_perc)
)

burden_palRenter <- colorNumeric(
  palette = "Blues",
  domain = sf_burden_pumas %>% 
    filter(
      status == "Renter"
    ) %>% 
    pull(burdened_30_perc)
)

floorarea_pal <- colorBin(
  palette = "Greens",
  bins = c(0,1000,5000,10000,max(my_parcel_zoning$unused_floorarea, na.rm = T))
)

units_pal <- colorBin(
  palette = "Oranges",
  bins = c(0,1,2,3) 
)

stories_pal <- colorBin(
  palette = "Blues",
  bins = c(0,1,2,3) 
)

ufa = round(sum(my_parcel_zoning$unused_floorarea),0)
mfa = round(sum(my_parcel_zoning$max_floorarea),0)

uu = round(sum(my_parcel_zoning$unused_units),0)
mu = round(sum(my_parcel_zoning$max_units),0)

us = round(sum(my_parcel_zoning$unused_stories),0)
ms = round(sum(my_parcel_zoning$max_stories),0)

```

Renter v. Owner Burdened Households
===================================== 


Column
-------------------------------------

Here is a map of the % of home owner populations that are burdened (spend over 30% of their annual income on mortgage and other related housing expenses) for each PUMA within San Francisco

### Burdened Owner Population

```{r}
sf_burden_pumas %>% 
  filter(
    status =="Owner"
  ) %>% 
  leaflet() %>% 
  addMapboxTiles(
    style_id = "streets-v11",
    username = "mapbox"
  ) %>% 
  addPolygons(
    fillColor = ~burden_palOwner(burdened_30_perc),
    fillOpacity = 0.5,
    color = "white",
    weight = 0.5,
    label = ~paste0(round(burdened_30_perc*100), "% of households paying 30%+ of income on housing"),
    highlightOptions = highlightOptions(
      weight = 2
    )
  ) %>% 
  addLegend(
    pal = burden_palOwner,
    values = ~burdened_30_perc,
    title = "% Cost-burdened<br>households"
  )
```

Column
-------------------------------------

Here is a map of the % of home renter populations that are burdened (spend over 30% of their annual income on rent and other related housing expenses) for each PUMA within San Francisco

### Burdened Renter Population

```{r}
sf_burden_pumas %>% 
  filter(
    status =="Renter"
  ) %>% 
  leaflet() %>% 
  addMapboxTiles(
    style_id = "streets-v11",
    username = "mapbox"
  ) %>% 
  addPolygons(
    fillColor = ~burden_palRenter(burdened_30_perc),
    fillOpacity = 0.5,
    color = "white",
    weight = 0.5,
    label = ~paste0(round(burdened_30_perc*100), "% of households paying 30%+ of income on housing"),
    highlightOptions = highlightOptions(
      weight = 2
    )
  ) %>% 
  addLegend(
    pal = burden_palRenter,
    values = ~burdened_30_perc,
    title = "% Cost-burdened<br>households"
  )
```

Parcel Metrics for Potential Density
===================================== 

Column
-------------------------------------

### Unused Floor Area

```{r}
gauge(
  ufa, 
  min = 0, 
  max = mfa, 
  gaugeSectors(
      success = c(0, mfa*.33 -1), 
      warning = c(mfa*.33, mfa*.66 - 1), 
      danger = c(mfa*.66, mfa)
  )
)
```

### Unused Units

```{r}
gauge(
  uu, 
  min = 0, 
  max = mu, 
  gaugeSectors(
      success = c(0, mu*.33 -1), 
      warning = c(mu*.33, mu*.66 - 1), 
      danger = c(mu*.66, mu)
  )
)
```

### Unused Stories

```{r}
gauge(
  us, 
  min = 0, 
  max = ms, 
  gaugeSectors(
      success = c(0, ms*.33 -1), 
      warning = c(ms*.33, ms*.66 - 1), 
      danger = c(ms*.66, ms)
  )
)
```

Column {.tabset}
-------------------------------------

### Unused Floor Area
Map of unused floor area by parcel. Two of the highest underutilized parcels are currently undergoing major renovations, so they will potentially be less underutilized in the future.

```{r}
leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolygons(
    data = my_parcel_zoning,
    fillColor = ~floorarea_pal(unused_floorarea),
    fillOpacity = 0.75,
    color = "white",
    weight = 0.5,
    label = ~round(unused_floorarea)
  ) %>% 
  addLegend(
    data = my_parcel_zoning,
    pal = floorarea_pal,
    values = ~unused_floorarea,
    title = "Unused Floor Area"
  )
```


### Additional Allowed Units

Map of additional units allowed on each parcel. No parcel has potential for more than 3 units, but these numbers are still quite high considering most of these buildings are not allowed to have more than 3 units based on zoning. The area consists of mostly of large homes, many of which have been subdivided into multiple units. This often results in units being larger than necessary since it is often easiest to make one entire floor into it's own unit. The southern section of this parcel contains many homes which have not been subdivided.

```{r}
leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolygons(
    data = my_parcel_zoning,
    fillColor = ~units_pal(unused_units),
    fillOpacity = 0.75,
    color = "white",
    weight = 0.5,
    label = ~round(unused_units)
  ) %>% 
  addLegend(
    data = my_parcel_zoning,
    pal = units_pal,
    values = ~unused_units,
    title = "Additional Units<br>Allowed"
  )
```

### Unused Stories 

Map of unused stories by parcel. This assumes each story would be 11 ft high. Therefore, even though the Residential Housing zoning category maximum height is 35 feet while the Residential Mixed zoning can be 40 ft high, buildings in both categories would be capped at 3 stories. However, Residential Mixed zoned buildings may realistically be able to include another story. 

```{r}
leaflet() %>% 
  addMapboxTiles(
    style_id = "light-v9",
    username = "mapbox"
  ) %>% 
  addPolygons(
    data = my_parcel_zoning,
    fillColor = ~stories_pal(unused_stories),
    fillOpacity = 0.75,
    color = "white",
    weight = 0.5,
    label = ~round(unused_stories)
  ) %>% 
  addLegend(
    data = my_parcel_zoning,
    pal = stories_pal,
    values = ~unused_stories,
    title = "Additional Stories<br>Allowed"
  )
```

